GeoSPARQL 1.1: an almost decadal update to the most important geospatial LOD standard

### 2021-04-21 manuscript.tex
7 pages

\institute{
    {
    SURROUND Australia Pty Ltd., Australia \&\\
    Australian National University, Australia\\
    \email{nicholas.car@surroundaustralia.com}\\
    }
    \and
    {
        Mainz University Of Applied Sciences, Germany\\
        \email{timo.homburg@hs-mainz.de}
    }
}

@@

\begin{abstract}
In 2012 the Open Geospatial Consortium published GeoSPARQL 
defining ``SPARQL extension functions'', ``RIF rules'' and ``an RDF/OWL ontology for [spatial] information''.\\

In the 8+ years since its publication, GeoSPARQL has become the most important spatial Semantic 
Web standard, as judged by references to it in other Semantic Web standards and its wide use in 
Semantic Web data.\\

An update to the standard was proposed in 2019 to deliver GeoSPARQL 1.1 in 2021 with a charter to: 
handle outstanding change requests and source new ones from the user community and 
to ``better present'' the standard, that is to better link all the standard’s parts and better 
document \& exemplify elements. Expected updates included alignments to other ontologies, 
handling of new spatial referencing systems, new geometry representations, and new artifact 
presentation.\\

@@

\section{Introduction}\label{sec:introduction}
The GeoSPARQL standard, issued in 2012 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
is one of, if not the most, popular \textit{Semantic Web} standards for spatial data.\footnote{References to GeoSPARQL in other well-known standards, 
such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) and CIDOC-CRM \cite{doerr2013crmgeo} suggests it is popular, as do the incoming links in Linked Open Vocabularies \url{https://lov.linkeddata.es/dataset/lov/vocabs/gsp} and the list of implementers that includes most of the popular triplestore vendors, a list of which has been compiled here: \url{https://github.com/opengeospatial/ogc-geosparql/issues/59}}

@@

In this publication we discuss the motivations behind updating GeoSPARQL 1.0 in \Cref{sec:motivation}, the content of the planned updates for the upcoming GeoSPARQL 1.1 release in \Cref{sec:newfeatures} and finally in \Cref{sec:conclusions} provide an outlook to further feature requests which are likely to be tackled in future GeoSPARQL 1.2 and 2.0 releases.

@@

The SWG's charter - its final scope of work - is also published by the OGC~\cite{abhayaratna2020ogc} and this guides 
the SWG's activities. Specific actions of the SWG and their staging are explained through the use of a publicly-available 
online task tracking system within the SWG's working online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql/projects/1}}.

At a high-level, proposed updates to GeoSPARQL by both the SDWWG and the SWG may be categorised as one of the following:

\begin{itemize}
    \item new geometry serializations
    \begin{itemize}
        \item[$-$] GeoJSON, KML and other popular formats
    \end{itemize} 
    \item new ontology classes to cater for more nuanced spatial information
    \item more spatial functions
    \begin{itemize}
        \item[$-$] implementing functions well-known in non \texttt{Semantic Web} spatial systems
    \end{itemize} 
    \item scalar spatial properties 
    \begin{itemize}
        \item[$-$] area, volume etc. alongside geometries
    \end{itemize} 
    \item better handling of Spatial (Coordinate) Reference Systems (SRS)
    \begin{itemize}
        \item[$-$] potentially allowing for automated coordinate serialization conversions
    \end{itemize} 
    \item Internet protocol-based selection of different geometries for features
\end{itemize}

Some of these proposed updates were predicted in GeoSPARQL 1.0, with the \textit{Future Work} section listing several of the 
points above as expected or potential.

The SWG's \textit{Charter}, anticipating that the more obvious updates such as new geometry serializations would certainly
be implemented, listed the following areas of investigation that emerged from SWG proponent's discussions:

\begin{itemize}
    \item a revision of the ``upper ontology'' structuring of GeoSPARQL
    \begin{itemize}
        \item[$-$] better/differently defining how GeoSPARQL's \texttt{Feature} and other classes relate to one another and to other fundamental concepts in the ontology
    \end{itemize} 
    \item alignments to other ontologies, perhaps \textit{W3C Time Ontology in OWL}~\cite{simon_cox_time_2017}
    \item catering for very different SRSes, such as Discrete Global Grid Systems
\end{itemize}

@@

\begin{itemize}
    \item \textbf{1.1}: extensions that are fully compatible with GeoSPARQL 1.0
    \item \textbf{1.2}: fully or mostly compatible extensions but which are larger additions to the standard's conceptual coverage
    \item \textbf{2.0}: a future GeoSPARQL that might be quite different and partly incompatible with GeoSPARQL 1.0
\end{itemize}

@@

An originally unforseen area of updates to GeoSPARQL is new modes of standard presentation. Motivatied by conceptual work within the W3C and the OGC for  
multi-part standards presentation, this has resulted in profile declarations, explained 
in the next section.

\section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
So far (as of April, 2021) the GeoSPARQL SWG has triaged change requests for GeoSPARQL releases and has addressed
many 1.1 requests which are the only ones we report here and which are accessible through the living document for the GeoSPARQL 1.1 standard \cite{ogcgeosparql11}. The next sections foreshadow likely 1.2 and 2.0 updates.

\subsection{Profile Declaration}\label{sec:profiledec}
One of the first actions undertaken by the SWG was to link the GeoSPARQL 1.0 elements through a \textit{profile} 
declaration, where a profile is a special type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}~\cite{atkinson_profiles_2020}. 
The specific motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
of which were easy to discover. As a result, some GeoSPARQL users were unaware of some of the resources and some
resources were accidentally duplicated or partly re-implemented. Profile declarations of this sort are anticipated, by the OGC, 
as being the \textit{best practice} way for it to deliver multi-part standards.

The profile declaration for GeoSPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
updated GeoSPARQL 1.0 resources. The profile declaration for GeoSPARQL 1.1 will be published at the same time as the 
1.1 releases, currently expected in mid-2021. All 1.0 and 1.1 release resources are currently available in draft form 
in the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}. The 1.1 releases' resources are:

\begin{enumerate}
    \item a profile declaration
    \item a specification document
    \item an RDF/OWL ontology document
    \item a Functions \& Rules vocabulary, derived from the ontology
    \item a Simple Features feature types vocabulary
    \item a set of RIF~\cite{kifer2013rif} rules
    \item SHACL~\cite{knublauch_shapes_2017} shapes for RDF data validation 
\end{enumerate}

@@

\begin{enumerate}
    \item \textbf{GeoJSON} (Geo- JavaScript Object Notation)~\cite{butler2016geojson}
    \item \textbf{KML} (Keyhole Markup Language)~\cite{nolan2014keyhole} 
    \item \textbf{DGGS} (Discrete Global Grid System)~\cite{sahr1998discrete}
\end{enumerate}

@@

GeoJSON \& KML have been much anticipated and were requested by the SDWWG and many users of 
GeoSPARQL, due to those formats' popularity.

The DGGS format is more forward-looking in that it is not driven by user demand but by predicted demand.
DGGS does not have a single, concrete format standard as the others do, nor is it ever likely to - different DGGSes will 
likely implement very different data formats - so GeoSAPRQL 1.1 makes generalized provisions for DGGS serializations but 
presents no detailed requirements for them, only stating that the specific DGGS must be identified.

@@

In addition to \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} defined in GeoSPARQL 1.0 
for use within SPARQL \texttt{FILTER} operations, 1.1 defines \href{http://www.opengis.net/def/function/geosparql/union2}{\emph{geof:union2}} and 
as well as \href{http://www.opengis.net/def/function/geosparql/boundingCircle}{\emph{geof:boundingCircle}}, \href{http://www.opengis.net/def/function/geosparql/centroid}{\emph{geof:centroid}}, \href{http://www.opengis.net/def/function/geosparql/concatLines}{\emph{geof:ConcatLines}} - concatenating a set of overlapping linestrings 
that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} gives an example 
of one of these new functions in use.

% \todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions. Then I will update the namespaces here. I would find it nice to also include the functions we have discussed yesterday: transformation of SRS and in between literal formats. Should we mention it here as future work or should we make a draft pull request for that so that people have a reference? I think the latter}
\small
\begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
# returns the circle geometry bounding all the geometries 
# of Feature <x>
SELECT (geof:BoundingCircle(?geo) AS ?circ)
WHERE {
    <x> geo:hasGeometry/geo:asWKT ?geo .
}
\end{lstlisting}
\normalsize

@@

Coincident with GeoSPARQL 1.1 development, the OGC API Features standard \cite{ogcapifeaturespart3} is being developed. It also proposes functions for filtering feature collections. While OGC API Features proposes the use of the Common Query Language (CQL) for filtering, it is also open to other query language implementations such as GeoSPARQL. When comparing the filter capabilities of CQL to GeoSPARQL, one can observe that the two query languages provide comparable spatial functionality, however the CQL proposed also supports spatiotemporal operators, which may be an addition to GeoSPARQL to be further explored in its contiuous development process.

\subsection{Ontology extensions}\label{sec:ontexts}
GeoSPARQL 1.1 - see Figure~\ref{fig:geosparql11ontology} for an overview - extends the GeoSPARQL ontology by adding a new class, \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{\texttt{geo:SpatialMeasure}}. This class represents a spatial measurement 
such as a volume, length, or area associated with a measurement amount and a unit of measure. It is the range of three newly-defined properties:
\href{http://www.opengis.net/ont/geosparql#hasArea}{\texttt{geo:hasArea}}, \href{http://www.opengis.net/ont/geosparql#hasLength}{\texttt{geo:hasLength}} and \href{http://www.opengis.net/ont/geosparql#hasVolume}{\texttt{geo:hasVolume}} which make these attributes of a geometry better accessible using 
SPARQL.

These additions address requests from the SDWWG \& SWG but
do more too: they open up GeoSPARQL to general patterns of measurement present in ontologies 
such as the W3C's \textit{SOSA}~\cite{haller_semantic_2017}. Similarly, 
the 1.1 release addition of property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, allows declarations of a geometry's SRS, independent of serializations and paves the way 
for future definition of SRSes in RDF, anticipated for GeoSPARQL 2.0.

>>>>

\institute{
    {
    %SURROUND Australia Pty Ltd., Australia \&\\
    Australian National University, Australia\\
    \email{nicholas.car@surroundaustralia.com}\\
    }
    \and
    {
        Mainz University Of Applied Sciences, Germany\\
        \email{timo.homburg@hs-mainz.de}
    }
}

@@

\begin{abstract}
In 2012 the Open Geospatial Consortium published GeoSPARQL 
defining ``SPARQL extension functions'', ``RIF rules'' and ``an RDF/OWL ontology for [spatial] information''.
In the 8+ years since its publication, GeoSPARQL has become the most important spatial Semantic 
Web standard, as judged by references to it in other Semantic Web standards and its wide use in 
Semantic Web data.\\

An update to GeoSPARQL was proposed in 2019 to deliver  1.1 in 2021 with a charter to: 
handle outstanding change requests and source new ones from the user community and 
to ``better present'' the standard, that is to better link all the standard’s parts and better 
document \& exemplify elements. Expected updates included alignments to other ontologies, 
handling of new spatial referencing systems, new geometry representations, and new artifact 
presentation.\\

@@

\section{Introduction}\label{sec:introduction}
The GeoSPARQL standard, issued in 2012 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
is one of the most popular \textit{Semantic Web} standards for spatial data.\footnote{References to GeoSPARQL in other well-known standards, 
such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) and CIDOC-CRM \cite{doerr2013crmgeo} suggests it is popular, as do the incoming links in Linked Open Vocabularies \url{https://lov.linkeddata.es/dataset/lov/vocabs/gsp} and the list of implementers that includes most of the popular triplestore vendors, a list of which has been compiled here: \url{https://github.com/opengeospatial/ogc-geosparql/issues/59}}

@@

Here we discuss the motivations behind updating GeoSPARQL 1.0 in \Cref{sec:motivation}, content of the planned GeoSPARQL 1.1 release in \Cref{sec:newfeatures} and finally in \Cref{sec:conclusions} provide an outlook to further feature requests which are likely to be tackled in future GeoSPARQL 1.2 and 2.0 releases.

@@

The SWG's charter - final scope of work - is also published by the OGC~\cite{abhayaratna2020ogc} and this guides 
the SWG's activities. Specific actions of the SWG and their staging are explained through the use of a publicly-available 
online task tracking system within the SWG's working online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql/projects/1}}.
At a high-level, proposed updates to GeoSPARQL by both the SDWWG and the SWG may be categorised as one of the following:

\begin{itemize}
    \item new geometry serializations
    \begin{itemize}
        \item GeoJSON, KML and other popular formats
    \end{itemize} 
    \item new ontology classes to cater for more nuanced spatial information
    \item more spatial functions
    \begin{itemize}
        \item implementing functions well-known in non \texttt{Semantic Web} spatial systems
    \end{itemize} 
    \item scalar spatial properties (area, volume etc. alongside geometries)
    \item better handling of Spatial (Coordinate) Reference Systems (SRS)
    \begin{itemize}
        \item potentially allowing for automated coordinate serialization conversions
    \end{itemize} 
    \item Internet protocol-based selection of different geometries for features
\end{itemize}

Some of these proposed updates were predicted in GeoSPARQL 1.0, with the \textit{Future Work} section listing several of the 
points above as expected or potential.
The SWG's \textit{Charter}, anticipating that the more obvious updates such as new geometry serializations would certainly
be implemented, listed the following areas of investigation that emerged from SWG proponent's discussions:

\begin{itemize}
    \item revising ``upper ontology'' GeoSPARQL structure - how its classes relate to fundamental concepts in ontology
    \item alignments to other ontologies, perhaps \textit{W3C Time Ontology in OWL}~\cite{simon_cox_time_2017}
    \item catering for very different SRSes, such as Discrete Global Grid Systems
\end{itemize}

@@

\begin{itemize}
    \item \textbf{1.1}: extensions that are fully compatible with GeoSPARQL 1.0
    \item \textbf{1.2}: fully or mostly compatible extensions but which are larger additions to the standard's conceptual coverage
    \item \textbf{2.0}: future GeoSPARQL likely incompatible with GeoSPARQL 1.0
\end{itemize}

@@

Originally unexpected, an area of updates to standard presentation. Motivatied by conceptual work within the W3C and OGC for  
multi-part standards, this has resulted in profile declarations explained in the next section.

\section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
So far (April, 2021) the GeoSPARQL SWG has triaged change requests for GeoSPARQL and has addressed
many 1.1 requests - the only ones we report here and which are accessible through the living document for the GeoSPARQL 1.1 standard \cite{ogcgeosparql11}. The next sections foreshadow likely 1.2 and 2.0 updates.

\subsection{Profile Declaration}\label{sec:profiledec}
One of the first SWG actions was to link GeoSPARQL 1.0 elements through a \textit{profile} 
declaration, where a profile is a type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}~\cite{atkinson_profiles_2020}. 
Motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
of which were easy to discover, so some GeoSPARQL users were unaware of some resources and some
resources were accidentally duplicated or partly re-implemented. Profile declarations of this sort are anticipated, by the OGC, 
as being the \textit{best practice} way for multi-part standards delivery.
The profile declaration for GeoSPARQL 1.0 will be published as a stand-alone resource along with some 
updated GeoSPARQL 1.0 resources and at the same time as the 
1.1 releases, currently expected in mid-2021. All 1.0 and 1.1 release resources are currently available in draft form 
in the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}. The 1.1 releases' resources are:

\begin{enumerate}
    \item a profile declaration
    \item a specification document
    \item an RDF/OWL ontology document
    \item a Functions \& Rules vocabulary, derived from the ontology
    \item a Simple Features feature types vocabulary
    \item a set of \textit{Rules Interchange Format} rules
    \item SHACL~\cite{knublauch_shapes_2017} shapes for RDF data validation 
\end{enumerate}

@@

\begin{enumerate}
    \item \textbf{GeoJSON} (Geo-JavaScript Object Notation)\footnote{\url{https://geojson.org}}%\cite{butler2016geojson}
    \item \textbf{KML} (Keyhole Markup Language)%\footnote{\url{http://doi.org/10.1007/978-1-4614-7900-0\_17}}%~\cite{nolan2014keyhole} 
    \item \textbf{DGGS} (Discrete Global Grid System)~\cite{sahr1998discrete}
\end{enumerate}

@@

GeoJSON \& KML have been much anticipated and were requested by the SDWWG and many users of 
GeoSPARQL, due to those formats' popularity. The DGGS format is more forward-looking in that it is not driven by user demand but by predicted demand.
DGGS does not have a single, concrete format standard as the others do, nor is it ever likely to - different DGGSes will 
likely implement very different data formats - so GeoSAPRQL 1.1 makes generalized provisions for DGGS serializations but 
presents no detailed requirements for them, only stating that the specific DGGS must be identified.

@@

In addition to \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} defined in GeoSPARQL 1.0 
for use within SPARQL \texttt{FILTER} operations, 1.1 defines \href{http://www.opengis.net/def/function/geosparql/union2}{\emph{geof:union2}} and 
as well as \href{http://www.opengis.net/def/function/geosparql/boundingCircle}{\emph{geof:boundingCircle}}, \href{http://www.opengis.net/def/function/geosparql/centroid}{\emph{geof:centroid}}, \href{http://www.opengis.net/def/function/geosparql/concatLines}{\emph{geof:ConcatLines}} - concatenating a set of overlapping linestrings 
that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} shows one new function in use.

% \todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions. Then I will update the namespaces here. I would find it nice to also include the functions we have discussed yesterday: transformation of SRS and in between literal formats. Should we mention it here as future work or should we make a draft pull request for that so that people have a reference? I think the latter}
\small
\begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
# returns circle bounding all geometries of Feature <x>
SELECT (geof:BoundingCircle(?geo) AS ?circ)
WHERE {<x> geo:hasGeometry/geo:asWKT ?geo .}
\end{lstlisting}
\normalsize

@@

Coincident with GeoSPARQL 1.1 development, the OGC API Features standard \cite{ogcapifeaturespart3} is being developed that proposes feature collections functions filtering. 
While it proposes the use of the Common Query Language (CQL) for filtering, it is also open to other query language implementations such as GeoSPARQL. When comparing the 
filter capabilities of CQL to GeoSPARQL, one can observe that the two query languages provide comparable spatial functionality, however the CQL proposed supports 
spatiotemporal operators, which may be an addition to GeoSPARQL to be further explored in its contiuous development process.

\subsection{Ontology extensions}\label{sec:ontexts}
GeoSPARQL 1.1 - see Figure~\ref{fig:geosparql11ontology} for an overview - extends the GeoSPARQL ontology by adding a new class, \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{\texttt{geo:SpatialMeasure}}. This class represents a spatial measurement 
such as a volume, length, or area associated with a measurement amount and a unit of measure. It is the range of three newly-defined properties:
\href{http://www.opengis.net/ont/geosparql#hasArea}{\texttt{geo:hasArea}}, \href{http://www.opengis.net/ont/geosparql#hasLength}{\texttt{geo:hasLength}} and \href{http://www.opengis.net/ont/geosparql#hasVolume}{\texttt{geo:hasVolume}} which make these attributes of a geometry better accessible using 
SPARQL. These additions address requests from the SDWWG \& SWG but
also open up GeoSPARQL to general patterns of measurement present in ontologies 
such as the W3C's \textit{SOSA}~\cite{haller_semantic_2017}. Similarly, 
the 1.1 release addition of property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, allows declarations of a geometry's SRS, independent of serializations and paves the way 
for future definition of SRSes in RDF, anticipated for GeoSPARQL 2.0.
